---
title: "Exercise 6"
subtitle: "Advanced Methods for Regression and Classification"
author: "Rita Selimi"
date: "27/11/2024"
output: 
  pdf_document:
    latex_engine: xelatex
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Set global options for all code chunks
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, tidy = TRUE)
```

# Predict Status

```{r load-data}
# Load the Loan dataset
data("Loan", package = "ROCit")

# View the dataset's structure
str(Loan)
head(Loan)
summary(Loan)
```

1.  **`Amount`**: The loan amount in thousands
2.  **`Term`**: The term of the loan in months
3.  **`IntRate`**: The interest rate as a decimal
4.  **`ILR`**: An index value representing the individual's loan rate
5.  **`EmpLen`**: Employment length categorized into 5 levels
6.  **`Home`**: Home ownership status categorized into 3 levels
7.  **`Income`**: The individual's income in dollars
8.  **`Status`**: Loan repayment status with two levels
9.  **`Score`**: A score associated with the individual’s loan profile

## 1. Linear Discrimant Analysis (LDA)

### (a) Apply lda()

LDA (Linear Discriminant Analysis) is a classification method that finds
the linear combination of predictors that best separates two or more
classes. It assumes the data for each class is normally distributed and
uses these linear combinations to assign new observations to one of the
classes.

#### Preprocessing

```{r}
# Ensure the response variable is a factor
Loan$Status <- as.factor(Loan$Status)

# Remove the constant variable "Term"
Loan <- Loan[, !(names(Loan) %in% "Term")]
```

The response variable Status must be a factor because discriminant
analysis methods (lda(), qda(), etc.) require categorical group labels
for classification tasks.

The column Term has the same value for all observations within at least
one group of Status. Since it doesn't vary, it provides no information
to distinguish between classes. Constant variables cause an error in
lda() because they make it impossible to calculate the covariance
matrix.

**Check for Missing Values**

```{r check-missing}
# Check for missing values
sum(is.na(Loan))
```

**Check for skewness and log transform highly skewed variables**

```{r}
# Load the necessary package
library(moments)

# Select numeric variables from the dataset
numeric_vars <- Loan[, sapply(Loan, is.numeric)]

# Compute skewness for each numeric variable
skewness_values <- sapply(numeric_vars, skewness)

# Display skewness values
skewness_values
```

```{r}
Loan$Amount <- log(Loan$Amount + 1)
Loan$Income <- log(Loan$Income + 1)  # Add 1 to avoid issues with zero values
```

**Scale Numerical Variables**

```{r scale-numerical}
numeric_vars <- sapply(Loan, is.numeric)
Loan[, numeric_vars] <- scale(Loan[, numeric_vars])

# Confirm scaling
summary(Loan)
```

LDA uses a covariance matrix, which is sensitive to the scale of the
predictors. If one variable has a much larger range than another (e.g.,
Income vs. IntRate), it will dominate the covariance calculation.
Scaling ensures that all variables contribute equally.

**Encoding Categorical Variables**

```{r}
Loan$EmpLen <- as.numeric(Loan$EmpLen)
Loan$Home <- as.numeric(Loan$Home)
```

**Checking Correlation**

```{r}
library(caret)

# Exclude `Status` from correlation analysis
numeric_vars <- Loan[, sapply(Loan, is.numeric) & names(Loan) != "Status"]

# Compute correlation matrix for numeric predictors
cor_matrix <- cor(numeric_vars)

# Find highly correlated features (cutoff = 0.9)
high_corr <- findCorrelation(cor_matrix, cutoff = 0.9)

# Remove highly correlated features from Loan dataset
Loan <- Loan[, -high_corr]
```

We are removing highly correlated predictors like `IntRate` to improve
the stability of the LDA model by avoiding redundancy in the covariance
matrix. The response variable `Status` is excluded from correlation
analysis to ensure only predictors are evaluated.

#### Split the data

```{r split-data}
# Set a seed for reproducibility
set.seed(12332281)

# Split data into training and test sets
train_indices <- sample(1:nrow(Loan), size = 2/3 * nrow(Loan))
train_data <- Loan[train_indices, ]
test_data <- Loan[-train_indices, ]
```

#### Apply lda()

```{r}
# Load required library
library(MASS)

# Apply Linear Discriminant Analysis
lda_model <- lda(Status ~ ., data = train_data)

# Print LDA model details
lda_model
```

1.  **Prior Probabilities:**\
    The model sees that the `FP` group dominates the training data, with
    about 87% of observations, while `CO` makes up only 13%.

2.  **Group Means:**\
    These are the average values of the predictors (`Amount`, `ILR`,
    etc.) for each class (`CO` and `FP`). For example:

    -   `ILR` is higher for `CO` (0.564) compared to `FP` (-0.061),
        which may help the model differentiate between the two groups.

3.  **Coefficients of Linear Discriminants:**\
    These coefficients show how much each predictor contributes to
    separating the two groups. For example:

    -   `ILR` has the largest coefficient (-0.853), making it the most
        important predictor for distinguishing `CO` from `FP`.
    -   Other predictors like `Income` (0.474) also contribute but to a
        lesser extent.
    -   Predictors with smaller coefficients, like `Amount` (-0.011),
        have little influence on classification.

### b) Compute the evaluation measures

```{r}
# Predict on the training data
lda_predictions <- predict(lda_model, train_data)$class

# Create a confusion matrix
confusion_matrix <- table(Predicted = lda_predictions, Actual = train_data$Status)
print(confusion_matrix)

# Calculate misclassification rate
misclassified <- sum(confusion_matrix) - sum(diag(confusion_matrix))  # FP + FN
total <- sum(confusion_matrix)  # FP + TN + FN + TP
misclassification_rate <- misclassified / total
cat("Misclassification Rate:", round(misclassification_rate, 4), "\n")

# Compute TPR (True Positive Rate) for each class
true_positives_fp <- confusion_matrix["FP", "FP"]
true_positives_co <- confusion_matrix["CO", "CO"]
actual_fp <- sum(confusion_matrix[, "FP"])
actual_co <- sum(confusion_matrix[, "CO"])

tpr_fp <- true_positives_fp / actual_fp  # TPR for FP
tpr_co <- true_positives_co / actual_co  # TPR for CO

# Compute TNR (True Negative Rate) for each class
false_positives_fp <- confusion_matrix["FP", "CO"]
false_positives_co <- confusion_matrix["CO", "FP"]
true_negatives_fp <- confusion_matrix["CO", "CO"]
true_negatives_co <- confusion_matrix["FP", "FP"]

tnr_fp <- true_negatives_fp / (true_negatives_fp + false_positives_co)  # TNR for FP
tnr_co <- true_negatives_co / (true_negatives_co + false_positives_fp)  # TNR for CO

# Compute balanced accuracy
balanced_accuracy <- (tpr_fp + tnr_fp) / 2
cat("Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
```

**Confusion Matrix:** `CO` is rarely predicted correctly (only **2 out
of 77** true `CO` instances are classified as `CO`). `FP` is predicted
correctly almost perfectly (**522 out of 523**).

3.  **Misclassification Rate:**
    -   Calculates the proportion of misclassified observations: $$
        \text{Misclassification Rate} = \frac{\text{FP} + \text{FN}}{\text{Total Observations}}
        $$The model misclassifies about 12.67% of the observations,
        indicating it performs well overall but struggles with minority
        class `CO`.
4.  **Balanced Accuracy:**
    -   First, it computes:
        -   **TPR:** The proportion of correctly classified instances of
            a given class.
        -   **TNR:** The proportion of correctly identified negatives
            for a given class.
    -   Then, averages these rates: $$
        \text{Balanced Accuracy} = \frac{\text{TPR} + \text{TNR}}{2}
        $$This value reflects the model's poor performance on the
        minority class (`CO`), as it heavily favors the majority class
        (`FP`).

The model performs well overall, with a low misclassification rate of
12.67%. The balanced accuracy of 83.24% indicates that the model
performs reasonably well, though it still struggles with the minority
class (`CO`). While it correctly identifies nearly all `FP` instances,
only 2 out of 77 `CO` instances are correctly classified. To further
improve the classification of the minority class, techniques like
oversampling, undersampling, or SMOTE could be considered.

### c) Predict for test data

```{r}
# Predict on the test data
test_predictions <- predict(lda_model, test_data)$class

# Create a confusion matrix
test_confusion_matrix <- table(Predicted = test_predictions, Actual = test_data$Status)
print(test_confusion_matrix)

# Calculate misclassification rate
test_misclassification_rate <- (test_confusion_matrix["CO", "FP"] + test_confusion_matrix["FP", "CO"]) / sum(test_confusion_matrix)
cat("Misclassification Rate (Test):", round(test_misclassification_rate, 4), "\n")

# Compute TPR (True Positive Rate) for each class
true_positives_fp_test <- test_confusion_matrix["FP", "FP"]
true_positives_co_test <- test_confusion_matrix["CO", "CO"]
actual_fp_test <- sum(test_confusion_matrix[, "FP"])
actual_co_test <- sum(test_confusion_matrix[, "CO"])

tpr_fp_test <- true_positives_fp_test / actual_fp_test
tpr_co_test <- true_positives_co_test / actual_co_test

# Compute TNR (True Negative Rate) for each class
false_positives_fp_test <- test_confusion_matrix["FP", "CO"]
false_positives_co_test <- test_confusion_matrix["CO", "FP"]
tnr_co_test <- test_confusion_matrix["FP", "FP"] / (test_confusion_matrix["FP", "FP"] + false_positives_fp_test)
tnr_fp_test <- test_confusion_matrix["CO", "CO"] / (test_confusion_matrix["CO", "CO"] + false_positives_co_test)

# Compute balanced accuracy
balanced_accuracy_test <- (tpr_co_test + tnr_co_test) / 2
cat("Balanced Accuracy (Test):", round(balanced_accuracy_test, 4), "\n")
```

The model performs poorly on the test data, with a misclassification
rate of 18.67% and a low balanced accuracy of 40.94%. While it correctly
classifies most `FP` instances (244 out of 246), it completely fails to
classify the minority class `CO`, misclassifying all 54 `CO` instances
as `FP`. This indicates significant bias toward the majority class and
poor generalization for the minority class.

## 2. Maximize the balanced accuracy

### a) Undersampling

Reduces the number of observations in the majority class by randomly
selecting a smaller subset, so it matches the size of the minority
class.

```{r}
library(dplyr)

# Count observations in each group
table(train_data$Status)

# Identify the smaller group size
min_count <- min(table(train_data$Status))

# Perform undersampling
set.seed(12332281)  # For reproducibility
train_data_undersampled <- train_data %>%
  group_by(Status) %>%
  sample_n(min_count) %>%
  ungroup()

# Check the new class balance
table(train_data_undersampled$Status)
```

#### Apply lda()

```{r}
# Apply Linear Discriminant Analysis
lda_model_undersampling <- lda(Status ~ ., data = train_data_undersampled)

# Print LDA model details
lda_model_undersampling
```

#### Compute the evaluation measures

```{r}
# Predict on the training data
lda_predictions_undersampling <- predict(lda_model_undersampling, train_data_undersampled)$class

# Create a confusion matrix
confusion_matrix <- table(Predicted = lda_predictions_undersampling, Actual = train_data_undersampled$Status)
print(confusion_matrix)

# Calculate misclassification rate
misclassified <- sum(confusion_matrix) - sum(diag(confusion_matrix))  # FP + FN
total <- sum(confusion_matrix)  # FP + TN + FN + TP
misclassification_rate <- misclassified / total
cat("Misclassification Rate:", round(misclassification_rate, 4), "\n")

# Compute TPR (True Positive Rate) for each class
true_positives_fp <- confusion_matrix["FP", "FP"]
true_positives_co <- confusion_matrix["CO", "CO"]
actual_fp <- sum(confusion_matrix[, "FP"])
actual_co <- sum(confusion_matrix[, "CO"])

tpr_fp <- true_positives_fp / actual_fp  # TPR for FP
tpr_co <- true_positives_co / actual_co  # TPR for CO

# Compute TNR (True Negative Rate) for each class
false_positives_fp <- confusion_matrix["FP", "CO"]
false_positives_co <- confusion_matrix["CO", "FP"]
true_negatives_fp <- confusion_matrix["CO", "CO"]
true_negatives_co <- confusion_matrix["FP", "FP"]

tnr_fp <- true_negatives_fp / (true_negatives_fp + false_positives_co)  # TNR for FP
tnr_co <- true_negatives_co / (true_negatives_co + false_positives_fp)  # TNR for CO

# Compute balanced accuracy
balanced_accuracy <- (tpr_fp + tnr_fp) / 2
cat("Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
```

#### Predict for test data

```{r}
# Predict on the test data
test_predictions_undersampling <- predict(lda_model_undersampling, test_data)$class

# Create a confusion matrix
test_confusion_matrix <- table(Predicted = test_predictions_undersampling, Actual = test_data$Status)
print(test_confusion_matrix)

# Calculate misclassification rate
test_misclassification_rate <- (test_confusion_matrix["CO", "FP"] + test_confusion_matrix["FP", "CO"]) / sum(test_confusion_matrix)
cat("Misclassification Rate (Test):", round(test_misclassification_rate, 4), "\n")

# Compute TPR (True Positive Rate) for each class
true_positives_fp_test <- test_confusion_matrix["FP", "FP"]
true_positives_co_test <- test_confusion_matrix["CO", "CO"]
actual_fp_test <- sum(test_confusion_matrix[, "FP"])
actual_co_test <- sum(test_confusion_matrix[, "CO"])

tpr_fp_test <- true_positives_fp_test / actual_fp_test
tpr_co_test <- true_positives_co_test / actual_co_test

# Compute TNR (True Negative Rate) for each class
false_positives_fp_test <- test_confusion_matrix["FP", "CO"]
false_positives_co_test <- test_confusion_matrix["CO", "FP"]
tnr_co_test <- test_confusion_matrix["FP", "FP"] / (test_confusion_matrix["FP", "FP"] + false_positives_fp_test)
tnr_fp_test <- test_confusion_matrix["CO", "CO"] / (test_confusion_matrix["CO", "CO"] + false_positives_co_test)

# Compute balanced accuracy
balanced_accuracy_test <- (tpr_co_test + tnr_co_test) / 2
cat("Balanced Accuracy (Test):", round(balanced_accuracy_test, 4), "\n")
```

### b) Oversampling

Increases the number of observations in the minority class by
duplicating samples (with replacement), so it matches the size of the
majority class.

```{r}
# Count observations in each group
table(train_data$Status)

# Identify the larger group size
max_count <- max(table(train_data$Status))

# Perform oversampling
train_data_oversampling <- train_data %>%
  group_by(Status) %>%
  sample_n(max_count, replace = TRUE) %>%
  ungroup()

# Check the new class balance
table(train_data_oversampling$Status)
```

#### Apply lda()

```{r}
# Apply Linear Discriminant Analysis
lda_model_oversampling <- lda(Status ~ ., data = train_data_oversampling)

# Print LDA model details
lda_model_oversampling
```

#### Compute the evaluation measures

```{r}
# Predict on the training data
lda_predictions_oversampling <- predict(lda_model_oversampling, train_data_oversampling)$class

# Create a confusion matrix
confusion_matrix <- table(Predicted = lda_predictions_oversampling, Actual = train_data_oversampling$Status)
print(confusion_matrix)

# Calculate misclassification rate
misclassified <- sum(confusion_matrix) - sum(diag(confusion_matrix))  # FP + FN
total <- sum(confusion_matrix)  # FP + TN + FN + TP
misclassification_rate <- misclassified / total
cat("Misclassification Rate:", round(misclassification_rate, 4), "\n")

# Compute TPR (True Positive Rate) for each class
true_positives_fp <- confusion_matrix["FP", "FP"]
true_positives_co <- confusion_matrix["CO", "CO"]
actual_fp <- sum(confusion_matrix[, "FP"])
actual_co <- sum(confusion_matrix[, "CO"])

tpr_fp <- true_positives_fp / actual_fp  # TPR for FP
tpr_co <- true_positives_co / actual_co  # TPR for CO

# Compute TNR (True Negative Rate) for each class
false_positives_fp <- confusion_matrix["FP", "CO"]
false_positives_co <- confusion_matrix["CO", "FP"]
true_negatives_fp <- confusion_matrix["CO", "CO"]
true_negatives_co <- confusion_matrix["FP", "FP"]

tnr_fp <- true_negatives_fp / (true_negatives_fp + false_positives_co)  # TNR for FP
tnr_co <- true_negatives_co / (true_negatives_co + false_positives_fp)  # TNR for CO

# Compute balanced accuracy
balanced_accuracy <- (tpr_fp + tnr_fp) / 2
cat("Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
```

#### Predict for test data

```{r}
# Predict on the test data
test_predictions_oversampling <- predict(lda_model_oversampling, test_data)$class

# Create a confusion matrix
test_confusion_matrix <- table(Predicted = test_predictions_oversampling, Actual = test_data$Status)
print(test_confusion_matrix)

# Calculate misclassification rate
test_misclassification_rate <- (test_confusion_matrix["CO", "FP"] + test_confusion_matrix["FP", "CO"]) / sum(test_confusion_matrix)
cat("Misclassification Rate (Test):", round(test_misclassification_rate, 4), "\n")

# Compute TPR (True Positive Rate) for each class
true_positives_fp_test <- test_confusion_matrix["FP", "FP"]
true_positives_co_test <- test_confusion_matrix["CO", "CO"]
actual_fp_test <- sum(test_confusion_matrix[, "FP"])
actual_co_test <- sum(test_confusion_matrix[, "CO"])

tpr_fp_test <- true_positives_fp_test / actual_fp_test
tpr_co_test <- true_positives_co_test / actual_co_test

# Compute TNR (True Negative Rate) for each class
false_positives_fp_test <- test_confusion_matrix["FP", "CO"]
false_positives_co_test <- test_confusion_matrix["CO", "FP"]
tnr_co_test <- test_confusion_matrix["FP", "FP"] / (test_confusion_matrix["FP", "FP"] + false_positives_fp_test)
tnr_fp_test <- test_confusion_matrix["CO", "CO"] / (test_confusion_matrix["CO", "CO"] + false_positives_co_test)

# Compute balanced accuracy
balanced_accuracy_test <- (tpr_co_test + tnr_co_test) / 2
cat("Balanced Accuracy (Test):", round(balanced_accuracy_test, 4), "\n")
```

**Training Data**

| Metric                 | **Undersampling** | **Oversampling** | **Better Strategy** |
|-------------------|------------------|------------------|------------------|
| Misclassification Rate | 35.71%            | 37.09%           | **Undersampling**   |
| Balanced Accuracy      | 67.27%            | 64.24%           | **Undersampling**   |

-   **Undersampling:** Achieves a slightly better misclassification rate
    and higher balanced accuracy on the training data.

**Test Data**

| Metric                 | **Undersampling** | **Oversampling** | **Better Strategy** |
|-------------------|------------------|------------------|------------------|
| Misclassification Rate | 37.33%            | 40.67%           | **Undersampling**   |
| Balanced Accuracy      | 67.82%            | 61.86%           | **Undersampling**   |

-   **Undersampling:** Performs better on test data, with both a lower
    misclassification rate and higher balanced accuracy.

**Undersampling is the more successful strategy.** It achieves better
results on both the training and test datasets, with a higher balanced
accuracy and a lower misclassification rate. While undersampling risks
losing information from the majority class, it avoids overfitting,
making it more effective for this dataset. Oversampling, on the other
hand, struggles with overfitting due to repeated samples in the minority
class, leading to poorer generalization on the test data.

## 3. Quadratic Discrimant Analysis (QDA)

Quadratic Discriminant Analysis (QDA) is a classification method that
models each class with its own covariance matrix, allowing for more
flexibility than LDA by capturing nonlinear decision boundaries.

### a) Undersampling

#### Apply qda()

```{r}
library(MASS)

# Train QDA on the undersampled training data
qda_model_undersampling <- qda(Status ~ ., data = train_data_undersampled)

# Print QDA model details
qda_model_undersampling
```

#### Compute the evaluation measures

```{r}
# Predict on the training data
qda_predictions_undersampling <- predict(qda_model_undersampling, train_data_undersampled)$class

# Create a confusion matrix
confusion_matrix <- table(Predicted = qda_predictions_undersampling, Actual = train_data_undersampled$Status)
print(confusion_matrix)

# Calculate misclassification rate
misclassified <- sum(confusion_matrix) - sum(diag(confusion_matrix))  # FP + FN
total <- sum(confusion_matrix)  # FP + TN + FN + TP
misclassification_rate <- misclassified / total
cat("Misclassification Rate:", round(misclassification_rate, 4), "\n")

# Compute TPR (True Positive Rate) for each class
true_positives_fp <- confusion_matrix["FP", "FP"]
true_positives_co <- confusion_matrix["CO", "CO"]
actual_fp <- sum(confusion_matrix[, "FP"])
actual_co <- sum(confusion_matrix[, "CO"])

tpr_fp <- true_positives_fp / actual_fp  # TPR for FP
tpr_co <- true_positives_co / actual_co  # TPR for CO

# Compute TNR (True Negative Rate) for each class
false_positives_fp <- confusion_matrix["FP", "CO"]
false_positives_co <- confusion_matrix["CO", "FP"]
true_negatives_fp <- confusion_matrix["CO", "CO"]
true_negatives_co <- confusion_matrix["FP", "FP"]

tnr_fp <- true_negatives_fp / (true_negatives_fp + false_positives_co)  # TNR for FP
tnr_co <- true_negatives_co / (true_negatives_co + false_positives_fp)  # TNR for CO

# Compute balanced accuracy
balanced_accuracy <- (tpr_fp + tnr_fp) / 2
cat("Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
```

#### Predict for test data

```{r}
# Predict on the test data
test_predictions_undercampling <- predict(qda_model_undersampling, test_data)$class

# Create a confusion matrix
test_confusion_matrix <- table(Predicted = test_predictions_undercampling, Actual = test_data$Status)
print(test_confusion_matrix)

# Calculate misclassification rate
test_misclassification_rate <- (test_confusion_matrix["CO", "FP"] + test_confusion_matrix["FP", "CO"]) / sum(test_confusion_matrix)
cat("Misclassification Rate (Test):", round(test_misclassification_rate, 4), "\n")

# Compute TPR (True Positive Rate) for each class
true_positives_fp_test <- test_confusion_matrix["FP", "FP"]
true_positives_co_test <- test_confusion_matrix["CO", "CO"]
actual_fp_test <- sum(test_confusion_matrix[, "FP"])
actual_co_test <- sum(test_confusion_matrix[, "CO"])

tpr_fp_test <- true_positives_fp_test / actual_fp_test
tpr_co_test <- true_positives_co_test / actual_co_test

# Compute TNR (True Negative Rate) for each class
false_positives_fp_test <- test_confusion_matrix["FP", "CO"]
false_positives_co_test <- test_confusion_matrix["CO", "FP"]
tnr_co_test <- test_confusion_matrix["FP", "FP"] / (test_confusion_matrix["FP", "FP"] + false_positives_fp_test)
tnr_fp_test <- test_confusion_matrix["CO", "CO"] / (test_confusion_matrix["CO", "CO"] + false_positives_co_test)

# Compute balanced accuracy
balanced_accuracy_test <- (tpr_co_test + tnr_co_test) / 2
cat("Balanced Accuracy (Test):", round(balanced_accuracy_test, 4), "\n")
```

### b) Oversampling

#### Apply qda()

```{r}
# Train QDA on the oversampling training data
qda_model_oversampling <- qda(Status ~ ., data = train_data_oversampling)

# Print QDA model details
qda_model_oversampling
```

#### Compute the evaluation measures

```{r}
# Predict on the training data
qda_predictions_oversampling <- predict(qda_model_oversampling, train_data_oversampling)$class

# Create a confusion matrix
confusion_matrix <- table(Predicted = qda_predictions_oversampling, Actual = train_data_oversampling$Status)
print(confusion_matrix)

# Calculate misclassification rate
misclassified <- sum(confusion_matrix) - sum(diag(confusion_matrix))  # FP + FN
total <- sum(confusion_matrix)  # FP + TN + FN + TP
misclassification_rate <- misclassified / total
cat("Misclassification Rate:", round(misclassification_rate, 4), "\n")

# Compute TPR (True Positive Rate) for each class
true_positives_fp <- confusion_matrix["FP", "FP"]
true_positives_co <- confusion_matrix["CO", "CO"]
actual_fp <- sum(confusion_matrix[, "FP"])
actual_co <- sum(confusion_matrix[, "CO"])

tpr_fp <- true_positives_fp / actual_fp  # TPR for FP
tpr_co <- true_positives_co / actual_co  # TPR for CO

# Compute TNR (True Negative Rate) for each class
false_positives_fp <- confusion_matrix["FP", "CO"]
false_positives_co <- confusion_matrix["CO", "FP"]
true_negatives_fp <- confusion_matrix["CO", "CO"]
true_negatives_co <- confusion_matrix["FP", "FP"]

tnr_fp <- true_negatives_fp / (true_negatives_fp + false_positives_co)  # TNR for FP
tnr_co <- true_negatives_co / (true_negatives_co + false_positives_fp)  # TNR for CO

# Compute balanced accuracy
balanced_accuracy <- (tpr_fp + tnr_fp) / 2
cat("Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
```

#### Predict for test data

```{r}
# Predict on the test data
test_predictions_oversampling <- predict(qda_model_oversampling, test_data)$class

# Create a confusion matrix
test_confusion_matrix <- table(Predicted = test_predictions_oversampling, Actual = test_data$Status)
print(test_confusion_matrix)

# Calculate misclassification rate
test_misclassification_rate <- (test_confusion_matrix["CO", "FP"] + test_confusion_matrix["FP", "CO"]) / sum(test_confusion_matrix)
cat("Misclassification Rate (Test):", round(test_misclassification_rate, 4), "\n")

# Compute TPR (True Positive Rate) for each class
true_positives_fp_test <- test_confusion_matrix["FP", "FP"]
true_positives_co_test <- test_confusion_matrix["CO", "CO"]
actual_fp_test <- sum(test_confusion_matrix[, "FP"])
actual_co_test <- sum(test_confusion_matrix[, "CO"])

tpr_fp_test <- true_positives_fp_test / actual_fp_test
tpr_co_test <- true_positives_co_test / actual_co_test

# Compute TNR (True Negative Rate) for each class
false_positives_fp_test <- test_confusion_matrix["FP", "CO"]
false_positives_co_test <- test_confusion_matrix["CO", "FP"]
tnr_co_test <- test_confusion_matrix["FP", "FP"] / (test_confusion_matrix["FP", "FP"] + false_positives_fp_test)
tnr_fp_test <- test_confusion_matrix["CO", "CO"] / (test_confusion_matrix["CO", "CO"] + false_positives_co_test)

# Compute balanced accuracy
balanced_accuracy_test <- (tpr_co_test + tnr_co_test) / 2
cat("Balanced Accuracy (Test):", round(balanced_accuracy_test, 4), "\n")
```

#### **Test Data**

| Metric                 | **Undersampling** | **Oversampling** | **Better Strategy** |
|-------------------|------------------|------------------|------------------|
| Misclassification Rate | 41.67%            | **40.67%**       | **Oversampling**    |
| Balanced Accuracy      | 67.29%            | **68.54%**       | **Oversampling**    |

-   **Oversampling:** Performs slightly better on the test set, with a
    lower misclassification rate and higher balanced accuracy. It
    generalizes better than undersampling in this case.

For QDA, **oversampling** performs slightly better than undersampling on
the test set, achieving a lower misclassification rate (40.67% vs.
41.67%) and a higher balanced accuracy (68.54% vs. 67.29%). This
suggests that oversampling helps the model generalize better to unseen
data by providing more examples of the minority class.

## 4. Regularized Discrimant Analysis (RDA)

Regularized Discriminant Analysis (RDA) is a classification method that
combines Linear Discriminant Analysis (LDA) and Quadratic Discriminant
Analysis (QDA) by introducing regularization parameters to balance
between the two, improving flexibility and stability for datasets with
multicollinearity or small sample sizes.

### a) Undersampling

#### Apply rda()

```{r}
library(klaR)

# Train RDA on the undersampled training data
rda_model_undersampling <- rda(Status ~ ., data = train_data_undersampled)

rda_model_undersampling
```

#### Predict for test data

```{r}
# Predict on the test data
test_predictions_undercampling <- predict(rda_model_undersampling, test_data)$class

# Create a confusion matrix
test_confusion_matrix <- table(Predicted = test_predictions_undercampling, Actual = test_data$Status)
print(test_confusion_matrix)

# Calculate misclassification rate
test_misclassification_rate <- (test_confusion_matrix["CO", "FP"] + test_confusion_matrix["FP", "CO"]) / sum(test_confusion_matrix)
cat("Misclassification Rate (Test):", round(test_misclassification_rate, 4), "\n")

# Compute TPR (True Positive Rate) for each class
true_positives_fp_test <- test_confusion_matrix["FP", "FP"]
true_positives_co_test <- test_confusion_matrix["CO", "CO"]
actual_fp_test <- sum(test_confusion_matrix[, "FP"])
actual_co_test <- sum(test_confusion_matrix[, "CO"])

tpr_fp_test <- true_positives_fp_test / actual_fp_test
tpr_co_test <- true_positives_co_test / actual_co_test

# Compute TNR (True Negative Rate) for each class
false_positives_fp_test <- test_confusion_matrix["FP", "CO"]
false_positives_co_test <- test_confusion_matrix["CO", "FP"]
tnr_co_test <- test_confusion_matrix["FP", "FP"] / (test_confusion_matrix["FP", "FP"] + false_positives_fp_test)
tnr_fp_test <- test_confusion_matrix["CO", "CO"] / (test_confusion_matrix["CO", "CO"] + false_positives_co_test)

# Compute balanced accuracy
balanced_accuracy_test <- (tpr_co_test + tnr_co_test) / 2
cat("Balanced Accuracy (Test):", round(balanced_accuracy_test, 4), "\n")
```

### b) Oversampling

#### Apply rda()

```{r}
rda_model_oversampling <- rda(Status ~ ., data = train_data_oversampling)

rda_model_oversampling
```

#### Predict for test data

```{r}
# Predict on the test data
test_predictions_oversampling <- predict(rda_model_oversampling, test_data)$class

# Create a confusion matrix
test_confusion_matrix <- table(Predicted = test_predictions_oversampling, Actual = test_data$Status)
print(test_confusion_matrix)

# Calculate misclassification rate
test_misclassification_rate <- (test_confusion_matrix["CO", "FP"] + test_confusion_matrix["FP", "CO"]) / sum(test_confusion_matrix)
cat("Misclassification Rate (Test):", round(test_misclassification_rate, 4), "\n")

# Compute TPR (True Positive Rate) for each class
true_positives_fp_test <- test_confusion_matrix["FP", "FP"]
true_positives_co_test <- test_confusion_matrix["CO", "CO"]
actual_fp_test <- sum(test_confusion_matrix[, "FP"])
actual_co_test <- sum(test_confusion_matrix[, "CO"])

tpr_fp_test <- true_positives_fp_test / actual_fp_test
tpr_co_test <- true_positives_co_test / actual_co_test

# Compute TNR (True Negative Rate) for each class
false_positives_fp_test <- test_confusion_matrix["FP", "CO"]
false_positives_co_test <- test_confusion_matrix["CO", "FP"]
tnr_co_test <- test_confusion_matrix["FP", "FP"] / (test_confusion_matrix["FP", "FP"] + false_positives_fp_test)
tnr_fp_test <- test_confusion_matrix["CO", "CO"] / (test_confusion_matrix["CO", "CO"] + false_positives_co_test)

# Compute balanced accuracy
balanced_accuracy_test <- (tpr_co_test + tnr_co_test) / 2
cat("Balanced Accuracy (Test):", round(balanced_accuracy_test, 4), "\n")
```

**Undersampling Results** - **Tuning Parameters (Gamma: 0.2258, Lambda:
0.0773):**\
- A **lower gamma** indicates that the model relies more on Quadratic
Discriminant Analysis (QDA), allowing for non-linear decision
boundaries.\
- A **lower lambda** suggests that the covariance matrices for the
classes are less regularized, meaning the model assumes more flexibility
in the spread of the data.

The model performs reasonably well, but the balanced accuracy suggests
it struggles slightly with classifying both classes equally.

**Oversampling Results** - **Tuning Parameters (Gamma: 0.4231, Lambda:
0.9488):**\
- A **higher gamma** indicates the model incorporates more Linear
Discriminant Analysis (LDA), favoring linear decision boundaries.\
- A **higher lambda** suggests strong regularization of the covariance
matrices, making the model more stable and less sensitive to outliers or
noise.

The model achieves higher balanced accuracy but has a slightly higher
misclassification rate.

-   **Oversampling** is preferable due to its higher balanced accuracy,
    which aligns with the goal of equal performance across both classes.
    The high lambda ensures model stability, while gamma's balance
    between LDA and QDA avoids overfitting to the minority class.
