---
title: "Exercise4_KaiThilenius_12423567"
author: "Kai Thilenius"
date: "2025-11-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction and appropriate split

Data set is a multivariate time series so we need a useful training/test split. To evaluate we used the RMSE in the last exercises already. We had a suitable train/test split with the chosen random seed so we just keep on using that one.

I don't think because it is a time series dataset we would need to use a time-based (chronological) split.

```{r, echo=TRUE}
 load("building.RData")
set.seed(69)
 n <- nrow(df)
 train <- sample(1:n, size = round(2/3 * n))
 test <- (1:n)[-train]
 head(df)
```

## 2. Partial Least Squares regression (PLS)

### 2 a) + b) Training PLS model and finding optimal components amount

We are using `plsr()` with 10 CV folds and scale the inputs. Also we cap the components before they explode radically agian with the `ncomp` from last exercise:

```{r, echo=TRUE}
library(pls)

# --- Fit PLS with 10-fold cross-validation ---
m_pls <- plsr(
  y ~ .,
  data       = df[train, ],
  scale      = TRUE,
  validation = "CV",
  segments   = 10,
  ncomp = 60 # from last exercise

)

# --- Extract CV RMSEs and find the optimal number of components ---
rmse_curve <- RMSEP(m_pls)$val[1, , ]        # RMSE for comps 0..K
opt_ncomp  <- max(1, which.min(rmse_curve) - 1)  # ensure >= 1
opt_rmse   <- rmse_curve[opt_ncomp + 1]      # matching CV RMSE

# --- Plot CV RMSE vs number of components, with annotation ---
validationplot(m_pls, val.type = "RMSEP", main = "PLS Cross-Validation (10-fold)")
# Add vertical and text annotation for the optimal point
abline(v = opt_ncomp, col = "red", lty = 2)
text(opt_ncomp + 1, opt_rmse,
     labels = paste0("opt = ", opt_ncomp, " comps\nRMSE = ", round(opt_rmse, 4)),
     pos = 4, col = "red", cex = 0.9)


```

The optimal number of components according to 10-fold cross-validation is *opt_ncomp* ,with a minimum cross-validated RMSE of *0.258*. The plot above illustrates this minimum (red dashed line). We do not use the RMSE + 1 StdErr to find a simpler model because we already have reduced the number of components a lot.

### 2 c) measured `y` values against the crossvalidated `y` values

```{r, echo=TRUE}
# --- Plot cross-validated predictions vs measured values ---
predplot(m_pls, ncomp = opt_ncomp,
         main = paste("Measured vs Cross-Validated Predictions (PLS, ncomp =", opt_ncomp, ")"),
         cex = 1, pch = 19, col = "steelblue")

# add 45° reference line
abline(0, 1, col = "red", lty = 2, lwd = 2)

```

We see a slight curve, which we also noticed with the models from previous exercises. Also visually in this scatterplot (first one we do actually) there seem to be no too big outliers and a quite equal distribution along the 45° line - confirming that our PLS model generalizes well on the trainig data under cross-validation.

### 2 d) predicted vs observed

```{r, echo=TRUE}
y_te    <- df$y[test]
yhat_te <- as.numeric(predict(m_pls, newdata = df[test, ], ncomp = opt_ncomp))

# RMSE on test
rmse_te_pls <- sqrt(mean((y_te - yhat_te)^2))
rmse_te_pls

# Predicted vs observed plot (TEST)
plot(y_te, yhat_te,
     xlab = "Observed y (TEST)",
     ylab = paste0("Predicted y (PLS, ncomp = ", opt_ncomp, ")"),
     main = paste0("PLS TEST: Predicted vs Observed (RMSE = ",
                   round(rmse_te_pls, 4), ")"),
     pch = 19)
abline(0, 1, lty = 2, col = "red", lwd = 2)

```

With *0.281* the prediction error to the test data with, our PLS model is not that much worse in test than in training (*0.258*, which makes a differece of only 0.023 RMSEPM).

When we compare it to the errors of previous models we stay in a similar area. PCR was with 0.265 a little better but also used way more components (33 instad of *8* with PLS).

## 3. Ridge Regression

### 3 a) `glmnet()` on training data - parameters for `lambda`?

```{r, echo=TRUE}
# --- Load packages and prepare data ---
library(glmnet)

Xtr <- as.matrix(df[train, setdiff(names(df), "y")])
ytr <- df$y[train]

# --- Fit Ridge regression (alpha = 0 → Ridge) ---
ridge_mod <- glmnet(
  Xtr, ytr,
  alpha = 0,         # Ridge penalty (L2)
  standardize = TRUE # predictors are standardized by default
)

# --- Plot the coefficient paths ---
plot(ridge_mod, xvar = "lambda", label = TRUE)
title("Ridge Regression Coefficient Paths", line = 2.5)

# --- Inspect the default lambda (lambda) sequence ---
lambda_values <- ridge_mod$lambda
cat("Number of lambda values:", length(lambda_values), "\n")
cat("Range of lambda values:\n")
print(range(lambda_values))

# --- Verify spacing: first 5 values (largest lambda → smallest lambda) ---
head(lambda_values)
tail(lambda_values)



```

Each colored line represents one predictors coefficient path as lambda changes. All of the 107 predictors are always there, never eliminated but just shrunken. The applied Ridge regression penalizes their size (magnitude) and pulls them closer to zero, without setting them exactly to zero. Moving right on x-axis actually corresponds to a smaller `lambda` (because of double negative)

Default `lambda` parameters used see output above.

The parameter `alpha` decides over the model used –\> 0 = Ridge (L2 penalty, no variable selection) and 1 = Lasso (L1 penalty) Regression.

### 3 b) optimal tuning parameters

```{r, echo=TRUE}
library(glmnet)

# --- Prepare training data ---
Xtr <- as.matrix(df[train, setdiff(names(df), "y")])
ytr <- df$y[train]

# --- Cross-validation for Ridge Regression (alpha = 0) ---
set.seed(69)  # reproducible folds
ridge_cv <- cv.glmnet(
  Xtr, ytr,
  alpha = 0,          # ridge penalty
  standardize = TRUE, # standardize predictors
  nfolds = 10          # 10-fold CV
)

# --- Plot cross-validation results ---
plot(ridge_cv)
title("Ridge Regression Cross-Validation", line = 2.5)

# --- Optimal lambda values ---
ridge_cv$lambda.min     # lambda that gives minimum CV error
ridge_cv$lambda.1se     # largest lambda within 1 SE of min (simpler model)

```

We use CV with the exact same fold and random seed as before. We take the lowest lambda and find the optimal tuning parameter by applying the -1se-rule we get a lambda of 0.163 — all predictors remain in the model but their magnitudes are reduced by this L2 penalty. The regression coefficients are listed here:

```{r, echo=TRUE}

coef_opt <- coef(ridge_cv, s = "lambda.min")
coef_opt

```

### 3 c) prediction

```{r, echo=TRUE}
library(glmnet)

# --- Prepare test data ---
Xte <- as.matrix(df[test, setdiff(names(df), "y")])
yte <- df$y[test]

# --- Predict with optimal model (lambda.min from cross-validation) ---
yhat_te_ridge <- as.numeric(predict(ridge_cv, newx = Xte, s = "lambda.min"))

# --- Compute RMSE on the test data ---
rmse_ridge_te <- sqrt(mean((yte - yhat_te_ridge)^2))
cat("Test RMSE (Ridge):", round(rmse_ridge_te, 4), "\n")

# --- Plot predicted vs observed test values ---
plot(yte, yhat_te_ridge,
     xlab = "Observed y (TEST)",
     ylab = "Predicted y (Ridge, lambda.min)",
     main = paste0("Ridge Regression: Predicted vs Observed (RMSE = ",
                   round(rmse_ridge_te, 4), ")"),
     pch = 19, col = "steelblue")
abline(0, 1, lty = 2, col = "red", lwd = 2)


```

Our model worked and predicted quite okay. Honestly i was expecting a better RMSESEM since we used all this fancy methods and graphs but are still in the same area as the models used in previous exercises with L0 being the best:

| Model                                  | RMSE on test data |
|----------------------------------------|-------------------|
| Principal component regression (PCR)   | 0.2653            |
| L0-norm                                | 0.2642            |
| Partial Least Squares regression (PLS) | 0.2809            |
| Ridge                                  | 0.2646            |

### 3 d) not strictly in/decreasing curves

I can also see this behaviour in at least two of the components (black on top, might be comp. 20 and orangy at the bottom, comp. 2). This might be a cause of connection inbetween different components - maybe because of the time series and various logs. When other components magnitude is decreaed, maybe there is new "space" for those two components to increase agian - agianst the normal assumption they might in/decrease continuously.
