---
title: "Exercise5_KaiThilenius_12423567"
author: "Kai Thilenius"
date: "2025-11-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 0. Importing the dataset - quite some struggle

The package "robstHD" is not available for my version of R since it seems to be no longer part of CRAN. I tried installing via GitHub but it failed too. I tired loading it from another package (ISLR2) which contains a "NCI60" too but this data doesn't match as it doesn't have any "protein" data. Installing package "robustHD" helped.

```{r, echo=TRUE}
data("nci60", package = "robustHD")
ls()
dim(gene); dim(protein)

```

-   59 samples (incomplete cases already removed)

-   22 283 gene features

-   162 protein features

```{r, echo=TRUE}
# choose a protein column with the largest variance (good spread)
j <- 130
cat("Using protein column:", j, "\n")

y <- as.numeric(protein[, j])   # response (length 64)
X <- as.matrix(gene)            # predictors (64 x 6830)

```

```{r, echo=TRUE}
set.seed(69)
n <- nrow(X)
train <- sample(seq_len(n), size = floor(0.75 * n))
test  <- setdiff(seq_len(n), train)

# build final matrices/vectors
Xtr <- X[train, , drop = FALSE]
ytr <- y[train]
Xte <- X[test,  , drop = FALSE]
yte <- y[test]

# quick sanity
length(train); length(test)
```

```{r, echo=TRUE}
library(glmnet)

# Perform ridge regression (alpha = 0)
set.seed(69)
ridge_cv <- cv.glmnet(
  Xtr, ytr,
  alpha = 0,
  nfolds = min(10, length(ytr)),
  standardize = TRUE,
  nlambda = 200,             # denser
  lambda.min.ratio = 1e-4    # explore much smaller lambda (default is 1e-3)
)

# --- Plot cross-validated MSE vs log(lambda) ---
plot(ridge_cv)
title(paste("Ridge Regression CV (Protein", j, ")"), line = 2.5)

# --- Extract optimal lambda values ---
ridge_cv$lambda.min    # minimum CV error
ridge_cv$lambda.1se    # simpler model within 1 std. error

# (optional) print them nicely
cat("Optimal lambda (min):", ridge_cv$lambda.min, "\n")
cat("1-SE lambda:", ridge_cv$lambda.1se, "\n")

```

We choose lambda as 319.2464 with best -1 SE.

## 1. Ridge regression

```{r, echo=TRUE}

# --- Extract the optimal lambda ---
lambda_opt <- ridge_cv$lambda.min
cat("Optimal lambda used:", lambda_opt, "\n")

# --- Extract the corresponding coefficients ---
ridge_coef <- coef(ridge_cv, s = lambda_opt)
ridge_coef[1:10, ]   # show the first 10 (intercept + top 9 genes)
cat("Number of predictors with nonzero coefficients:", sum(ridge_coef != 0) - 1, "\n")

# --- Predict on the test set ---
yhat_ridge <- as.numeric(predict(ridge_cv, newx = Xte, s = lambda_opt))

# --- Compute test RMSE ---
rmse_ridge <- sqrt(mean((yte - yhat_ridge)^2))
cat("Test RMSE (ridge):", round(rmse_ridge, 4), "\n")

# --- Plot predicted vs observed values on the test set ---
plot(yte, yhat_ridge,
     pch = 19, col = "steelblue",
     xlab = "Observed y (test)",
     ylab = paste0("Predicted y (ridge, lambda = ", round(lambda_opt, 2), ")"),
     main = paste0("Ridge Regression — Test Predictions (RMSE = ", round(rmse_ridge, 4), ")"))
abline(0, 1, col = "red", lty = 2, lwd = 2)



```

Ridge regression tends to “average” predictions when data dimensionality is extreme relative to sample size. This leads to lambda getting incredible big which reveals in an "empty model" like prediction. I tried to find a better solution but also with only allowing smaller lambdas we did not really get better predictions. So we have to keep it like this :(

## 2 Lasso regression

```{r, echo=TRUE}
set.seed(69)

# --- LASSO CV (alpha = 1) ---
# If your ridge CV chose very large lambdas earlier, give lasso a wider/smaller range:
lasso_cv <- cv.glmnet(
  Xtr, ytr,
  alpha = 1,                 # LASSO
  nfolds = min(10, length(ytr)),
  standardize = TRUE,
  nlambda = 200,
  lambda.min.ratio = 1e-4    # explore smaller lambda than the default (1e-3)
)

# CV curve
plot(lasso_cv); title("LASSO Cross-Validation (MSE vs log(lambda))", line = 2.5)

# Optimal lambdas
lambda_min <- lasso_cv$lambda.min
lambda_1se <- lasso_cv$lambda.1se
cat("lambda.min (min CV error):", lambda_min, "\n")
cat("lambda.1se (simpler):",     lambda_1se, "\n")
```

```{r, echo=TRUE}
# Predict on TEST with lambda.min
yhat_lasso <- as.numeric(predict(lasso_cv, newx = Xte, s = "lambda.min"))
rmse_lasso <- sqrt(mean((yte - yhat_lasso)^2))
cat("Test RMSE (LASSO, lambda.min):", round(rmse_lasso, 4), "\n")

# Plot predicted vs observed (TEST)
plot(yte, yhat_lasso,
     pch = 19, col = "tomato",
     xlab = "Observed y (TEST)",
     ylab = paste0("Predicted y (LASSO, lambda = ", signif(lambda_min, 3), ")"),
     main = paste0("LASSO — Test Predictions (RMSE = ", round(rmse_lasso, 4), ")"))
abline(0, 1, lty = 2, col = "gray30", lwd = 2)


```

## 3 Elastic net regression

```{r, echo=TRUE}

set.seed(69)

# --- 1) Define an alpha grid (you can play with this)
alpha_grid <- c(0, 0.25, 0.5, 0.75, 1)   # 0=ridge, 1=lasso

# Use a richer lambda search than defaults (helps with p >> n)
nl <- 200
lmr <- 1e-4

fits <- vector("list", length(alpha_grid))
summ <- data.frame(
  alpha = alpha_grid,
  lambda.min = NA_real_,
  lambda.1se = NA_real_,
  cvm.min    = NA_real_,
  nnz.min    = NA_integer_,  # non-zeros at lambda.min
  nnz.1se    = NA_integer_   # non-zeros at lambda.1se
)

for (i in seq_along(alpha_grid)) {
  a <- alpha_grid[i]
  cvfit <- cv.glmnet(
    Xtr, ytr,
    alpha = a,
    nfolds = min(10, length(ytr)),
    standardize = TRUE,
    nlambda = nl,
    lambda.min.ratio = lmr
  )
  fits[[i]] <- cvfit

  # record metrics
  summ$lambda.min[i] <- cvfit$lambda.min
  summ$lambda.1se[i] <- cvfit$lambda.1se
  summ$cvm.min[i]    <- min(cvfit$cvm)

  # count non-zero coefficients (exclude intercept)
  b.min <- as.matrix(coef(cvfit, s = "lambda.min"))[-1, , drop = FALSE]
  b.1se <- as.matrix(coef(cvfit, s = "lambda.1se"))[-1, , drop = FALSE]
  summ$nnz.min[i] <- sum(b.min != 0)
  summ$nnz.1se[i] <- sum(b.1se != 0)
}

summ

op <- par(mfrow = c(1,2))
plot(summ$alpha, summ$cvm.min, type="b", pch=19,
     xlab=expression(alpha), ylab="CV MSE (min)",
     main="Elastic Net: CV error vs alpha")
grid()

plot(summ$alpha, summ$nnz.min, type="b", pch=19, ylim = range(c(summ$nnz.min, summ$nnz.1se)),
     xlab=expression(alpha), ylab="# non-zero coefficients",
     main="Sparsity vs alpha")
lines(summ$alpha, summ$nnz.1se, type="b", pch=1)
legend("topright", c("lambda.min","lambda.1se"), pch=c(19,1), bty="n")
grid()
par(op)
```

```{r, echo=TRUE}
cand <- subset(summ, nnz.min >= 1)                 # avoid empty
best_idx <- which.min(cand$cvm.min)
best_row <- cand[best_idx, ]
best_row
# locate the stored cvfit for this alpha
i_alpha <- which(alpha_grid == best_row$alpha)
cvbest  <- fits[[i_alpha]]

lambda_use <- "lambda.min"   # or "lambda.1se" if you want simpler

# Coefficients
coef_best <- coef(cvbest, s = lambda_use)
nnz_best  <- sum(as.matrix(coef_best)[-1, ] != 0)

cat("Chosen alpha:", best_row$alpha, "\n")
cat("Chosen lambda:", if(lambda_use=="lambda.min") best_row$lambda.min else best_row$lambda.1se, "\n")
cat("# non-zero predictors:", nnz_best, "\n")

# Test predictions + RMSE
yhat_enet <- as.numeric(predict(cvbest, newx = Xte, s = lambda_use))
rmse_enet <- sqrt(mean((yte - yhat_enet)^2))
cat("Elastic Net TEST RMSE:", round(rmse_enet, 4), "\n")

# Plot predicted vs observed (TEST)
plot(yte, yhat_enet, pch=19, col="#2c7fb8",
     xlab="Observed y (TEST)",
     ylab=paste0("Predicted y (EN, alpha=", best_row$alpha, ", ", lambda_use, ")"),
     main=paste0("Elastic Net — Test (RMSE=", round(rmse_enet,4), ")\n",
                 "#non-zero=", nnz_best))
abline(0,1,lty=2,col="gray30",lwd=2)

```

## 4. Adaptive Lasso regression

```{r, echo=TRUE}
library(glmnet)
## 1) Ridge coefficients (exclude intercept)
beta_ridge <- as.numeric(coef(ridge_cv, s = "lambda.min"))[-1]

## 2) Adaptive weights  w_j = 1/(|beta_ridge_j|^gamma + eps)
##    Use a non-tiny eps to avoid monstrous weights from near-zero ridge betas.
gamma <- 1
eps   <- 1e-3
w     <- 1 / (abs(beta_ridge)^gamma + eps)

## (optional but recommended) cap extremely large weights
w <- pmin(w, quantile(w, 0.99))  # trim the top 1%

## 3) CV Adaptive Lasso
set.seed(69)
alasso_cv <- cv.glmnet(
  Xtr, ytr,
  alpha = 1,                    # L1 penalty
  penalty.factor = w,           # <-- adaptive weights
  standardize = TRUE,
  nfolds = min(10, length(ytr)),
  nlambda = 200,
  lambda.min.ratio = 1e-4       # explore smaller lambdas but not too tiny
)

plot(alasso_cv); title("Adaptive Lasso — CV", line = 2.5)

lambda_min_al <- alasso_cv$lambda.min
lambda_1se_al <- alasso_cv$lambda.1se
cat("lambda.min:", lambda_min_al, "  lambda.1se:", lambda_1se_al, "\n")

## 4) Use the more stable 1-SE model to avoid overfitting
coef_1se <- as.matrix(coef(alasso_cv, s = "lambda.1se"))
nnz_1se  <- sum(coef_1se[-1, 1] != 0)
cat("# non-zero (1-SE):", nnz_1se, "\n")

## 5) Test predictions + RMSE (1-SE)
yhat_te_al <- as.numeric(predict(alasso_cv, newx = Xte, s = "lambda.1se"))
rmse_te_al <- sqrt(mean((yte - yhat_te_al)^2))
cat("Adaptive Lasso TEST RMSE (1-SE):", round(rmse_te_al, 4), "\n")

plot(yte, yhat_te_al,
     pch = 19, col = "#1b9e77",
     xlab = "Observed y (TEST)",
     ylab = paste0("Predicted y (Adaptive Lasso, lambda.1se = ", signif(lambda_1se_al,3), ")"),
     main = paste0("Adaptive Lasso — Test (RMSE=", round(rmse_te_al, 4), ")\n",
                   "#non-zero = ", nnz_1se))
abline(0, 1, lty = 2, col = "gray30", lwd = 2)


```

Seems

## A - which model was useful

| Model                  | RMSE on test data |
|------------------------|-------------------|
| Ridge regression       | 0.3945            |
| Lasso regression       | 0.369             |
| Elastic net regression | 0.3873            |
| Adaptive Lasso         | 0.3498            |

All the models operate in a similar RMSE are. They are all not ideally or even suitable for predicting but Adaptive Lasso has the best. RMSE value on the test data for this one selected protein.

## B - most interpretable

Among the tested models, the **Lasso** and **Adaptive Lasso** provide the best interpretability because they perform **automatic variable selection** by setting many coefficients exactly to zero.\
This yields a sparse model that highlights which genes (predictors) are most relevant for predicting the protein response.

In contrast, **Ridge regression** keeps all coefficients non-zero, making it difficult to interpret which variables matter most, even though it may perform similarly in terms of RMSE.

The **Elastic Net** offers a good compromise: it keeps correlated genes together (useful when features are strongly correlated) and maintains moderate sparsity, making it both more stable than Lasso and more interpretable than Ridge.

Therefore, **for interpretability**, the *Lasso* (or *Adaptive Lasso*, if stable) is the most useful model, while Ridge is least interpretable but often most stable.

### Disclaimer

ChatGPT 5 might have been used to generate some of the code and texts.
